Community Detection Methods for the Generalized Random Dot Product Graph Model

### 2021-05-24 proposal/slides.Rmd
updated slides

---
title: Community Detection Methods for the Generalized Random Dot Product Graph Model
subtitle: Dissertation Proposal
author: John Koo
date: 'June 2021'
output: 
  beamer_presentation:
    fig_crop: no
    theme: 'default'
    colortheme: 'beaver'
    includes:
      in_header: page_headers.tex
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
- \setbeamertemplate{itemize items}[circle]
---

@@

# Introduction

@@

Example: Stochastic Block Model with two communities

@@

Want to estimate $X$ given $A$, or alternatively, 
interpoint distances, inner products, or angles.

@@

Example: Assortative SBM

@@

Example: SBM (cont'd)

* $A \sim BernoulliGraph(X X^\top)$
* $A \approx \hat{X} \hat{X}^\top$
  * $\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$
* Apply clustering algorithm (e.g., $K$-means) on $\hat{X}$

```{r, fig.height = 8, fig.width = 8, out.width = '50%'}
A.eigen <- eigen(A)
X.hat <- A.eigen$vectors[, 1:2] %*% diag(A.eigen$values[1:2] ** .5)
plot(X.hat, asp = 1, col = z, xlab = NA, ylab = NA)
```

@@

**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to 
$A \sim GRDPG_{3, 1}(X U)$ for block diagonal $X$ 
constructed from $\{\lambda^{(kl)}\}_2$ and 
predetermined $U \in \mathbb{O}(4)$.

@@

**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_K)$ is equivalent to 
$A \sim GRDPG_{p, q}(X U)$ such that

* $p = K (K + 1) / 2$
* $q = K (K - 1) / 2$
* $U$ is orthogonal and predetermined for each $K$
* $X$ is block diagonal and composed of $\{\lambda^{(kl)}\}_K$  
$\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
to different communities, then $x_i^\top x_j = 0$.

**Remark** (non-uniqueness of the latent configuration):  
$A \sim GRDPG_{p, q}(X U) \implies A \sim GRDPG_{p, q}(X U Q)$ 
$\forall Q \in \mathbb{O}(p, q)$

>>>>

---
title: Community Detection Methods for the Generalized Random Dot Product Graph Model
subtitle: Dissertation Proposal
author: John Koo
date: 'June 2021'
output: 
  beamer_presentation:
    fig_crop: no
    theme: 'default'
    colortheme: 'beaver'
    includes:
      in_header: page_headers.tex
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square,comma}
- \usepackage{verbatim}
- \usepackage{amsthm}
- \usepackage{comment}
---

@@

@@

Example 1: Stochastic Block Model with two communities

@@

Want to estimate $X$ from $A$, or alternatively, 
interpoint distances, inner products, or angles.

@@

Example 1 (cont'd): Assortative SBM ($p q > r^2$) with $K = 2$

@@

Example 1 (cont'd)

If we want to perform community detection,

1. Note that $A \sim BernoulliGraph(X X^\top)$ is a RDPG
2. Compute the ASE $A \approx \hat{X} \hat{X}^\top$  
$\hat{X} = \hat{V} \hat{\Lambda}^{1/2}$
3. Apply clustering algorithm (e.g., $K$-means) to $\hat{X}$

```{r, fig.height = 4, fig.width = 4, out.width = '50%'}
A.eigen <- eigen(A)
X.hat <- A.eigen$vectors[, 1:2] %*% diag(A.eigen$values[1:2] ** .5)
plot(X.hat, asp = 1, col = z, xlab = NA, ylab = NA,
     main = 'ASE of the adjacency matrix drawn from SBM')
```

@@

**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_2)$ is equivalent to 
$A \sim GRDPG_{3, 1}(X U)$ for block diagonal $X$ 
constructed from $\{\lambda^{(kl)}\}_2$ and $U \in \mathbb{O}(4)$.

@@

**Theorem** (KTT): $A \sim PABM(\{\lambda^{(kl)}\}_K)$ is equivalent to 
$A \sim GRDPG_{p, q}(X U)$ with

* $p = K (K + 1) / 2$
* $q = K (K - 1) / 2$
* $U \in \mathbb{O}(K^2)$
* $X$ is block diagonal and composed of $\{\lambda^{(kl)}\}_K$ 
with each block corresponding to a community  
$\implies$ if $x_i^\top$ and $x_j^\top$ are two rows of $X U$ corresponding 
to different communities, then $x_i^\top x_j = 0$.

**Remark** (non-uniqueness of the latent configuration):  
If $A \sim GRDPG_{p, q}(X)$, then $A \sim GRDPG_{p, q}(X Q)$ for any $Q$ in 
the indefinite orthogonal group with signature $p, q$.
